A Type System for Safe and Scalable Parallelism using Logical Regions

### 2012-07-05 doc/popl2013/langdef.tex

\begin{thm}
Let $\oton{e}$ be well-typed Legion expressions, each with its own privileges $\Phi_i$.
Let $M$ be a region mapping, $L$ be local value mapping,
$H$ be a heap typing, and $S$ be an initial state of the store satisfying the consistency properties:
$$
\begin{array}{l}
M \sim \Omega \\
L \sim_H \Gamma \\
S \sim H
\end{array}
$$
If $\Phi_i \#_M \Phi_j$ for $1 \leq i < j \leq n$, then any parallel execution of expressions
$\oton{e}$ will result in a valid interleaving of memory operations.
\end{thm}

@@

TODO: talk about atomic case

@@

Let $e_1$ and $e_2$ be expressions that require privileges $\Phi_1$ and
$\Phi_2$.  Now let $e_3$ be some subexpression of $e_1$ and $e_4$ be some 
subexpression of $e_2$, with corresponding privileges $\Phi_3$ and $\Phi_4$.

Proper scheduling of the evaluation of $e_3$ and $e_4$ depends on whether
they interfere.  If they do intefere (and must therefore be serialized), we
will have $E_3 \cancel\# E_4$, and by the above result, $\Phi_3 \cancel\# \Phi_4$.

We now show that this will also imply that $\Phi_1 \cancel\# \Phi_2$, and
unless sequential behavior has been relaxed by the programmer, the
evaluations of $e_1$ and $e_2$ will be serialized.  Since the evaluation of
$e_3$ is a subsequence of the evaluation of $e_1$ and the evaluation of $e_4$
is a subsequence of the evaluation of $e_2$, the evaluations of $e_3$ and $e_4$
are properly serialized as well.

Conversely, if $e_1$ and $e_2$ are judged to be noninterfering, the schedule
does not need to check for interference between $e_3$ and $e_4$, as their
noninterference is guaranteed.

Proof is straightforward once armed with the following lemmas:

\framebox{$\Phi_1 \#_M \Phi_2 \Leftrightarrow \Phi_1^* \#_M \Phi_2^*$}

\framebox{$\Phi_1 \#_M \Phi_2 \wedge \Phi_3 \subseteq \Phi_2 \Rightarrow \Phi_1 \#_M \Phi_3$}

>>>>

\begin{thm}
\label{thm:parallelexec}
Let $\oton{e}$ be well-typed Legion expressions, each with its own privileges $\Phi_i$.
Let $M$ be a region mapping, $L$ be local value mapping,
$H$ be a heap typing, and $S$ be an initial state of the store satisfying the consistency properties:
$$
\begin{array}{l}
M \sim \Omega \\
L \sim_H \Gamma \\
S \sim H
\end{array}
$$
If $\Phi_i \#_M \Phi_j$ for $1 \leq i < j \leq n$, then any parallel execution of expressions
$\oton{e}$ will result in a valid interleaving of memory operations.
\end{thm}

@@

\subsection{Relaxed Coherence}

The analysis above did not consider the coherence modes on each operation in the memory traces,
making the worst-case assumption that all were using the $excl$ mode.  However, if two or more
subexpressions contain calls to functions that use $atomic$ and/or $simult$ coherence modes
on some of their regions, options other than serial execution may be guaranteed to 
result in valid interleavings even if constituent traces have some locations in common.  There
are two cases of particular interest that we will cover.  In each case, the entirety of the 
reasoning from the previous section is repeated with minor modifications, so we will limit our
discussion to a general description of the differences rather than the complete proofs.

The first important case occurs when all apparently-interfering operations have coherence modes
of either $atomic$ or $simult$.  In this case, we can see that the conditions required for
coherence and sequential equivalence are satisfied for all locations in $L_{excl}$.  The
requirements for locations in $L_{atomic}$ allow for any permutation of the constituent traces,
so the runtime is free to serialize the execution of the subexpressions in any order, not just
the original program order.

In order to convert the required analysis to one that can be performed on regions, we modify
the noninterference operator for effects to consider only locations in $L_{excl}$ and show that
traces that a noninterfering under this modified check satisfy the requirement for atomic
execution.  We then modify the noninterference operator on privileges to allow 
$M(r_1) \cap M(r_2) \not= \emptyset$ if either $atomic(r_1)$ or $simult(r_1)$ is in $Q_1$ and
either $atomic(r_2)$ or $simult(r_2)$ is in $Q_2$, and can show that subexpressions with 
privileges that can be shown to be ``atomically noninterfering'' (under a dynamic region mapping
$M$) will always have effects that are compatible with atomically-reordered execution.  Finally,
in the special case where each subexpression has $atomic(r)$ or $simult(r)$ in $Q_i$ for every $r$
in $\Phi_i$, we can eliminate the dynamic check entirely.

The second case is a subset of the first, in which all apparently-interfering operations have
the $simult$ coherence mode.  In this case, all the coherence and sequential equivalence 
checks are trivially satisfied, and any interleaving of the constituent traces will be valid.
We again modify the noninterference criteria for effects, traces, and privileges, this time
only allowing $simult$ coherence annotations, and can show that subexpression with privileges
that can be shown to be ``simulataneously noninterfering'' (under mapping $M$) can be executed
in parallel.  This will include all the cases covered by Theorem~\ref{thm:parallelexec} as well
as those where the programmer has used the $simult$ coherence mode to explicitly allow parallel
execution.  Again, a special (but important) case exists where all privileges for each
subexpression are paired with $simult$ coherence mode declarations, allowing the safety of
parallel execution to be determined at compile time.

@@

The enforcement of a (locally) valid interleaving at every step in the operational semantics
captures the hierarchical nature of the scheduler used by the Legion runtime.  As the cost of
a scheduling decision is quadratic in the number of tasks (more precisely, the number of 
privileges for those tasks), it should be clear that a hierarchical scheduling algorithm will
be vastly more efficient than a global algorithm that must consider the privileges of every
concurrently executing expression.
However, a scheduling algorithm that makes decisions hierarchically must do so with limited
information, and must respect all required scheduling dependencies while ideally imposing as
few extra dependencies as possible.

Let $e_1$ and $e_2$ be well-typed Legion expressions that using privileges $\Phi_1$ and
$\Phi_2$.  Now let $e_3$ be some subexpression of $e_1$ and $e_4$ be some 
subexpression of $e_2$, with corresponding privileges $\Phi_3$ and $\Phi_4$.

By Theorem~\ref{thm:parallelexec}, if the runtime can verify $\Phi_1 \#_M \Phi_2$, then 
expressions $e_1$ and $e_2$ may be evaluated in parallel.  The evaluation of the subexpressions
$e_3$ and $e_4$ will therefore be allowed to run in parallel, and we must show that this is
permissible.  We do so by proving that any memory traces that can result from the evaluation
of the subexpression must be noninterfering (allowing parallel execution by
Lemma~\ref{lem:noninteffects}).

\begin{thm}
Let $e_1$ and $e_2$ be well-typed Legion expressions that using privileges $\Phi_1$ and
$\Phi_2$.  Now let $e_3$ be some subexpression of $e_1$ and $e_4$ be some 
subexpression of $e_2$.  For any memory traces $E_3$ and $E_4$ that result from evaluation of
$e_1$ and $e_2$ (with the usual consistent $M$, $L$, $H$, and $S$), $E_3$ and $E_4$ must be
noninterfering.
\end{thm}

\begin{proof}
Let $E_1$ be the result of an evaluation of $e_1$, with $E_3$ being the memory trace from the
evaluation of $e_3$ within the expression tree.  $E_1$ will be formed from a tree of valid
interleavings, each of which must include all the memory operations of each consitituent trace.
By induction over the number of intermediate subexpressions between $e_1$ and $e_3$, we
can show that any memory operation $\epsilon_3$ that is in $E_3$ must also be in $E_1$.
Similarly, any memory operation $\epsilon_4$ that is in $E_4$ must also be in $E_2$.  We have
$E_1 \# E_2$ from Lemma~\ref{lem:nonintpriv}, so and therefore $\epsilon_3 \# \epsilon_4$.  Since
this holds for all $\epsilon_3$ in $E_3$ and all $\epsilon_4$ in $E_4$, we have $E_3 \# E_4$.
\end{proof}

Although a decision that $e_1$ and $e_2$ are noninterfering guarantees that $e_3$ and $e_4$ are
noninterfering, the converse is not true.  It is possible for $e_1$ and $e_2$ to be judged to be
potentially interfering and be serialized in their entirety even when the subexpressions could
have been proven to be noninterfering.  This loss of precision can impact performance, and did
for a few times during the implementation of our experiments.  When the performance impact is
significant, two options exist to improve the precision.  In cases where the $e_1$ and $e_2$ are
in fact noninterfering, the partitioning of regions can be further refined to make the
disjointness visible in the privileges.  Alternately, and this is the only option when the 
interference is dynamically determined, the code can be refactored to life the subexpressions
(or some expression containing them) up to the same level, allowing the dynamic test based on
the region mapping to infer noninterference in more cases.  Although this can increase the
overhead of scheduling decisions, judicious use of this technique in practice has not shown
any measurable performance degradation.
